---
title: "BDA 1"
author: "Man Kit Yip"
date: "22/10/2017"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---
#1. Basic Statistics (1%)
###(a) Suppose you ask a group of 10 students at Birkbeck College how mnany brothers and sisters they have. The number obtained are as follows:

###2 3 0 5 2 1 1 0 3 3

###I will create a vector to hold this data on a number of siblings:
```{r}
siblings <- c(2, 3, 0, 5, 2, 1, 1, 0, 3, 3)
```

###Find the following measures of central tendency:

###(i) the mean,
```{r}
mean(siblings)
###The average number of siblings is 3.
```

###(ii) the median
```{r}
median(siblings)
###The median number of siblings is also 3.
```


###(iii) the mode
```{r}
mode <- function(v) {
  uniqv <- unique(siblings)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
results <-mode(siblings)
print(results)
###The most common number of siblings is 3.
```


###Find the following measures of spread:

###(iv) the variance
```{r}
var(siblings)
###The variance in the numnber of siblings is 2.444444.
```


###(v) the standard deviation
```{r}
sd(siblings)
###The Standard deviation in the set is 1.563472.
```

###(b) Suppose these 10 students have the following age:

###23 25 18 45 30 21 22 19 29 35

###(i) Find the covariance and correlation between the number of siblings and their age
```{r}
age <- c(23, 25, 18, 45, 30, 21, 22, 19, 29, 35)
cov(siblings, age)
cor(siblings, age)
###The covariance between the number of siblings and their age is 11.88889.
###The Correlation between the number of siblings and their age is 0.9116971.
```


###(ii) Is there a positive or negative or no correlation between the two?
```{r}
###As it can seen the correlation coefficient between siblings and age is 0.9116971 which is very close to +1, this suggests a strong linear association.
```


###(iii) Is there a causation between the two? Justify your answer.
```{r}
###However a correlation does not equates to causation. To put it into context, how old the childrens are, does not directly affect how many children each couple has in total. The correlation may be due to chance, or other unknown factors which may affect it e.g. psychological stress for couples to bring up another newborn baby so soon, and fertility of parents.
```


#2. Getting familiar with R (2%) [Textbook 2.10]
#This exercise involves the Boston housing data set.

###(a) To begin, load in the Boston data set. The Boston data set is part of the MASS library (MASS)
```{r}
library(MASS)
```

###Now the data set is contained in the object Boston. Use the following command to read about the data set:
```{r}
Boston
```

###You could get more info by the following command:
```{r}
?Boston
```

 Boston {MASS}	R Documentation
Housing Values in Suburbs of Boston

Description

The Boston data frame has 506 rows and 14 columns.

Usage

Boston
Format

This data frame contains the following columns:

crim
per capita crime rate by town.

zn
proportion of residential land zoned for lots over 25,000 sq.ft.

indus
proportion of non-retail business acres per town.

chas
Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).

nox
nitrogen oxides concentration (parts per 10 million).

rm
average number of rooms per dwelling.

age
proportion of owner-occupied units built prior to 1940.

dis
weighted mean of distances to five Boston employment centres.

rad
index of accessibility to radial highways.

tax
full-value property-tax rate per \$10,000.

ptratio
pupil-teacher ratio by town.

black
1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town.

lstat
lower status of the population (percent).

medv
median value of owner-occupied homes in \$1000s.

Source

Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81–102.

Belsley D.A., Kuh, E. and Welsch, R.E. (1980) Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley.


###How many rows are in this data set? The number of rows in the Boston data set is: 
```{r}
nrow(Boston)
###506
```


###How many columns? The number of columns is:
```{r}
ncol(Boston)
###14
```


###What do the rows and columns represent?
```{r}
### From the R documentation "Housing Values in Suburbs of Boston". I can see that the rows represents different neighbourhoods of the Boston area and the columns represents the different variables, including:
### 1) crim: per capita crime rate by town.
### 2) zn: proportion of residential land zoned for lots over 25,000 sq.ft.
### 3) indus: proportion of non-retail business acres per town.
### 4) chas: Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).
### 5) nox: nitrogen oxides concentration (parts per 10 million).
### 6) rm: average number of rooms per dwelling.
### 7) age: proportion of owner-occupied units built prior to 1940.
### 8) dis: weighted mean of distances to five Boston employment centres.
### 9) rad: index of accessibility to radial highways.
### 10) tax: full-value property-tax rate per \$10,000.
### 11) ptratio: pupil-teacher ratio by town.
### 12) black: 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town.
### 13) lstat: lower status of the population (percent).
### 14) medv: median value of owner-occupied homes in \$1000s.
```


###(b) Make some pairwise scatterplots of the predictors (columns) in this data set. Describe your findings. 
```{r}
### To see how strongly the the different variables correlate with one another, I will have to plot the pairwise data to generate a graphical representation and also run cor tests to show it numerically.
pairs(Boston, pch='.')
cor(Boston, use = "everything")
```

###(c) Are any of the predictors associated with per capita crime rate? If so, explain the relationship.
```{r}
###As it can be seen from the scatterplot (and the correlation coefficient), variable 1) crim represents shows a moderately strong positive correlation with variable 9) rad with a correlation coefficient of 0.62550515. This is closely followed by variable 10) tax with a correlation coefficient of ax 0.58276431. This means that as the crime rate per capita increases, the index of accessibility to radial highways tends to be higher. Also as the crime rate per capital increases, the full-value property-tax rate tends to be higher. Other predictors shows weaker correlations < +/-0.5 co-efficiency; with the exception of variable 4) chas which seems to have almost no correlation with variable 1) crim with a co-efficiency of -0.05589158 and a rather high p value of 0.2094. There appears to be negative a correlation between dis, black and medv with crim rate. 

cor.test(Boston$crim, Boston$rad)
cor.test(Boston$crim, Boston$tax)
cor.test(Boston$crim, Boston$chas)
```


###(d) Do any of the suburbs of Boston appear to have particularly high crime rates?
```{r}
Boston[order(Boston$crim),]
Boston[1]
### I have sorted the data in ascending order of crime rates, and their corresponding neighbourhood represented by a number i.e. 1 - 506. As it can be seen the highest crime rate is in neighbourhood 381 at 88.97620 per capita followed by 419, 406 and so on. Taken as a whole, the neighbourhoods denoted by 401, 404, 405, 406, 411, 414, 415, 418, 419 and 428 seems to have high crime rates. If these are indeed clusters within the same area, this would indicate that particular area has higher crime rates in comparison to the rest of the city of Boston.
```


###Tax rates?
```{r}
Boston[order(Boston$tax),]
Boston[order(10)]
### The neighbourhoods of 493 to 489 appears to have the highest full-value property-tax rate per \$10,000, due having the highest value property-tax value (at $711), this may indicate 'expensive' areas to live in. The neighbourhoods 357 to 488 appears to have the next highest full-value property-tax rate per \$10,000 (at $666). The neighbourhood wth the lowest full-value property-tax rate per \$10,000 is 354 (at $187) and the second lowest neighbourhoods 121 to 127 (at $188).
```


###Pupil-teacher ratios?
```{r}
Boston[order(Boston$ptratio),]
Boston[11]
###The neighbourhoods 355 and 356 has the highest pupil-teacher ratios at 22 students per teacher. This is closely followed by neighbourhoods 128 to 142 with the next highest pupilp-teacher ratios at 21.2 pupils per teacher. This means the schools in those neighbourhoods are more deprived and the average class size is larger, which may in turn affect the education of the pupils who went there from the local catchment area.
```

###Comment on the range of each predictor.
```{r}
summary(Boston)
plot(Boston$crim)
plot(Boston$tax)
plot(Boston$ptratio)
###The range from the lowest to the highest value varies widely for the three predictors: 
###1) crime rates: Min 0.00632 to Max 88.97620, mean = 3.61352; the increase from the 3rd Quartile to the max value is very steep, this means that there are a few neighbourhoods with disproportionately high crime rates. These could be outliers.
###10) tax rates: Min 187.0 to Max 711.0, mean = 408.2; this is a huge range which the maximum value over 3 times that of the lowest value. There could be outliers.
###11) pupil-teacher ratio; Min 12.60 to Max 22.00, mean = 18.46; the increase from the min value to the 1st Quartile is very steep, this means that there are a few neighbourhoods with disproportionately higher number of teachers for the pupils there. However there are no clear outliers.

#For full details, please see below...
```

###(e) How many of the suburbs in this data set bound the Charles river?
```{r}
sum(Boston$chas)
###As seen above 4) chas: Charles River dummy variable, a value of 1 indicates that the suburbs are bound by the river Charles, and a value 0 if otherwise. Since the numbers can only be 1 or 0, a sum of 35 means that there are 35 suburbs in this data set which bound the Charles river.
```


###(f) What is the median pupil-teacher ratio among the towns in this data set?
```{r}
median(Boston$ptratio)
###The median pupil-teacher ratio among the (of all of the) towns in this data set is about 19 students per teacher as shown below. this is different to the average pupil-teacher ratio which is slightly lower.
```


###(g) Which suburbs of Boston has lowest median value of owner occupied homes?
```{r}
Boston[order(Boston$medv),]
Boston[14]
### I have sorted the data in ascending order of the 14) medv: median value of owner-occupied homes in \$1000s. The following ten suburbs appears to have the lowest median value of owner occupied homes 399, 406, 401, 400, 415, 490, 386, 402, 416, 388 (ranges from $5000 to $7400). The next lowest median value of owner occupied homes are also 'numerically very similar'. Their low values may indicate that properties in the area are not very desirable. 
```

###What are the values of the other predictors for that suburb?
```{r}
Boston[ c(399,406), ]

###The suburb 399 and 406 both have the lowest median value of owner occupied homes. The other values of the other predictors are also shown below. 
###For subrub 399: crim = 38.3518, zn = 0, indus = 18.1, chas = 0, nox = 0.693, rm = 5.453, age = 100, dis = 1.4896, rad = 24, tax = 666, ptratio = 20.2, black = 396.90, lstat = 30.59, medv = 5.
###For suburb 406: crim = 67.9208, zn = 0, indus = 18.1, chas = 0, nox = 0.693, rm = 5.683, age = 100, dis = 1.4254, rad = 24, tax = 666, ptratio = 20.2, black = 384.97, lstat = 22.98, medv = 5.
```


###How do those values compare to the overall ranges for those predictors?
```{r}
Boston[range(399, 406),]
sapply(Boston,quantile)
###If you compare those values with the mean/ median in the summary above, or the ranges below, you can see that in both neighbourhoods: 
### 1) The crime rate is much higher than the 'average'.
### 2) There are no residential land zoned for lots over 25,000 sq.ft.
### 3) The proportion of non-retail business acres per town is relatively high.
### 4) They are both outside the bound of the Charles River.
### 5) The Nitrogen oxides concentration is relatively high.
### 6) The average number of rooms per dwelling are relatively low.
### 7) The proportion of owner-occupied units built prior to 1940 is extremely high.
### 8) The weighted mean of distances to five Boston employment centres is low.
### 9) The index of accessibility to radial highways is very high.
### 10) The full-value property-tax rate per \$10,000 is very high.
### 11) The pupil-teacher ratio by town is high.
### 12) The proportion of blacks by town is very high.
### 13) The percentage of lower status of the population is very high.
### 14) The median value of owner-occupied homes in \$1000s is extremely low.
```

###Comment on your findings.
```{r}
###From the results, I can see that in the areas of which crime rate is very high e.g. suburb 406:
###The average properties are small and very old; also there are no big mansions. 
###The neighbourhood has little business activity and there are many job centres for the umemployed. 
###It is in the outskirts of the city, and is next to multiple highways; the air is polluted with Nitrogen oxides (from cars).
###The tax rate on the properties is very high, whereas the properties' value itself is extremely low. It is mostly occupied by the lower status members of the society.
###The schools have larger class sizes, and the proportion of blacks in the area is very high.
```

###(h) In this data set, how many of the suburbs average more than seven rooms per dwelling? 
```{r}
Boston[(Boston[,6]>7),]
###As it can be seen there are a total of 64 suburbs which average more than seven rooms per dwellings. Those neighbourhoods can be found below...
```


###More than eight rooms per dwelling?
```{r}
Boston[(Boston[,6]>8),]
###As it can be seen there are a total of 13 suburbs which average more than seven rooms per dwellings. Those neighbourhoods can be found below...
```


###Comment on the suburbs that average more than eight rooms per dwelling.
```{r}
rbind(sapply(Boston[Boston$rm > 8,], mean), sapply(Boston, mean))
###If you compare the suburbs that average more than eight rooms per dwelling with all towns in Boston, in terms of their other variables (taken into consideration the range, average, and outliers etc), it be concluded that:
### 1) The crime rate is very low.
### 2) There is a higher proportion of residential land zoned for lots over 25,000 sq.ft.
### 3) The proportion of non-retail business acres per town ranges is generally low.
### 4) Most are within the bound of the Charles River while a few are outside.
### 9) The index of accessibility to radial highways is lower.
### 10) The full-value property-tax rate per \$10,000 is lower.
### 11) The pupil-teacher ratio by town is lower than average.
### 13) The percentage of lower status of the population is very low.
### 14) The median value of owner-occupied homes in \$1000s is very high.
```


#3. Linear Regression (2%) [Textbook 3.13 (a-f)]

#In this exercise you will create some simulated data and will fit simple linear regression models to it. Make sure to use set.seed(1) prior to starting part (a) to ensure consistent results.
```{r}
set.seed(1)
```


###(a) Usibng the rnorm() function, create a vector, x, containing 100 observations drawn from a N (0,1) distribution, i.e., a normal distribution with mean 0 and variance 1. This represents a feature, X.
```{r}
set.seed(1)
x <- rnorm(100, mean = 0, sd = sqrt(1))
###since the variance is 1, the squareroot of 1 is also 1, therefore the standard deviation is 1.
```


###(b) Using the rnorm() function, create a vector, eps, containing 100 observations drawn from a N (0,0.25) distribution i.e. a normal distribution with mean 0 and variance 0.25.
```{r}
eps<- rnorm(100, mean = 0, sd = sqrt(0.25))
###Since the4 variance is 0.25, the squareroot of that is 0.5, therefore the standard deviation is 0.5.
###Note that I do not need to set seed for eps.
```


###(c) Using x and eps, generate a vector y according to the model

Y = -1 + 0.5X + e.

```{r}
y <- -1 + 0.5*x + eps
###note that e. is epsilon
```

###What is the length of the vector y? What are the values of B0 and B1 in this linear model?
```{r}
length(y)
###The length of vector y is 100.
###In this linear model, B0 is -1 and B1 is 0.5.
```


###(d) Create a scattrerplot displaying the relationship between x and y. 
```{r}
plot(x, y, main = "A scatterplot showing the relationship between x and y")
###Please see the scatterplot below...
```


###Comment on what you observe.
```{r}
###The scatterplot shows that there seems to be a positive correlation between x and y. As x increases, y increases, vice versa.
```

###(e) Fit a least square linear model to predict y using x. Comment on the model obtained. How do ^B0 and ^B1 compare to B0 and B1?
```{r}
lm.fit <- lm(y ~ x)
summary(lm.fit)
###The Estimated ^B0 is -1.01885, and the estimated ^B1 is 0.49947.
###The Residual Standard Error is 0.4814 on 98 degrees of freedom. This implies that any prediction on the observed y based on x would be off by about 0.4814 (no unit) on average when using the least square line.
###It would still be fairly close to the actual values B0 and B1.
```

###(f)Display the least squares line on the scatterplot obtained in (d). Draw the population regression line on the plot, in a different colour. Use the legend() command to create to create an appropriate legend.
```{r}
plot(x,y, main = "Linear regression showing the relationship between x and y")
abline(-1, 0.5, col = "red")
abline(lm.fit, col = "blue")
legend(x = c(0, 3), y = c(-2.5, -2), legend = c("Population", "fitted model"), col = c("red", "blue"), lwd = 3)

###The red line represents the true relationship, AKA population regression line.
###The blue line represents the least square line; it is the least square estimate for the model based on the data.
```


###References:
Tutorialspointcom. 2017. Wwwtutorialspointcom. [Online].  Available from: https://www.tutorialspoint.com/r/r_mean_median_mode.htm. Last accessed 24 Oct 2017

Ajitesh. (2014). Learn R - How to Extract Rows & Columns from Data Frame. Available: https://dzone.com/articles/learn-r-how-extract-rows. Last accessed 26 Oct 2017.

Gareth james et al, . 2017. An introduction to statistical learning. [Online]. Available from: http://www-bcf.usc.edu/~gareth/ISL/code.html. Last accessed 28 Oct 2017

###Bibliography:
James, Gareth. (2017). Statistical learning. In: et al An introduction to Statistical learning. 8th ed. New York: springer. p15-52.

James, Gareth. (2017). Statistical learning. In: et al An introduction to Statistical learning. 8th ed. New York: springer. p59-120.

Ting Ting, Han. (2017). Introduction. Lecture delivered to Msc Data Science Big Data Analytics using R, level 7, 4th October 2017

Ting Ting, Han. (2017). Basic Statistics. Lecture delivered to Msc Data Science Big Data Analytics using R, level 7, 11th October 2017

Ting Ting, Han. (2017). Linear Regression. Lecture delivered to Msc Data Science Big Data Analytics using R, level 7, 18th October 2017

###End of Document